Supplementary Materials: Eco-evolutionary dynamics, density-dependent dispersal, and collective behavior: implications for salmon metapopulation robustness

### 2018-01-21 Manuscript/FinalDraft3/msv10.tex
labels to figure 4

% \Blindtext
%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Eco-evolutionary dynamics, density dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
\author{
Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Thilo Gross${}^3$, Peter A. H. Westley${}^{4}$, \& Jonathan W. Moore${}^{5}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$University of Bristol, Bristol, UK\\
${}^4$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@

Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
This diversity-stability linkage occurs when there are asynchronous population dynamics within the metapopulation. 
This asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}. 
For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population's coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}. 
Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.

@@

\begin{align}
  \label{eq:R}
  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} {\rm p}(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i.
\end{align}

\noindent The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it is unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

>>>>

% \Blindtext
%\title{Simple rules yield complex communities: deconstructed species interactions and the assembly of communities}
%\title{Community assembly and dynamics by the deconstruction of species interactions}
\title{Eco-evolutionary dynamics, density dependent dispersal, and collective behavior: implications for salmon metapopulation robustness}
\author{
Justin D. Yeakel${}^{1,2,*}$, Jean P. Gibert${}^{1}$, Thilo Gross${}^3$, Peter A. H. Westley${}^{4}$, \& Jonathan W. Moore${}^{5}$ \\
${}^1$School of Natural Sciences, University of California, Merced, Merced CA, USA \\
${}^2$The Santa Fe Institute, Santa Fe NM, USA \\
${}^3$Department of Engineering Mathematics, University of Bristol, Bristol, UK\\
${}^4$College of Fisheries and Ocean Sciences, University of Alaska, Fairbanks, Fairbanks AK, USA \\
${}^5$Earth${}_2$Oceans Research Group, Simon Fraser University, Burnaby BC, Canada \\
${}^*$To whom correspondence should be addressed: jdyeakel@gmail.com
}

@@

Intraspecific diversity can increase the resilience and stability of species or metapopulations. 
This diversity-stability linkage can arise when there are asynchronous population dynamics within the metapopulation. 
Such asynchrony will increase the potential for demographic rescue \citep{Brown:1977gk,Earn:2000fm} and also decrease the variability of processes that integrate across the metapopulation \citep{Anonymous:2015gf}. 
For example, different responses to climate variability within populations of a rare plant reduced fluctuations in abundance \citep{Abbott:2017hl}. 
This statistical buffer has traditionally been quantified as the Portfolio Effect (PE), which is the ratio of the population's coefficient of variation (CV) to the CV of the aggregated metapopulation \citep{Thibaut:2012km}. 
Strengthened portfolio effects are expected to increase the robustness of metapopulations to external disturbances, and by extension promote persistence \citep{Thibaut:2012km}.
In contrast, homogenization of populations leading to greater synchronization and weakened PE may be a harbinger of metapopulation collapse and extinction \citep{Carlson:2011ce}.

@@

\begin{align}
  \label{eq:R}
  &R_i[\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\theta_i] = \\ \nonumber
  &\int_{-\infty}^\infty r_{\rm max}\exp\left\{\frac{(x_i-\theta_i)^2}{2\tau^2}\right\} g(x_i,\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t),\sigma^2) {\rm d}x_i +\tilde{P}_i\\ \nonumber
  &= \frac{r_{\rm max} \tau  }{\sqrt{\sigma ^2+\tau ^2}}\exp\left\{-\frac{(\theta_i-(\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)))^2}{2 \left(\sigma ^2+\tau ^2\right)}\right\} +\tilde{P}_i,
\end{align}

\noindent where $g(x_i)$ is the Gaussian probability density function for the trait $x_i$. The mismatch between the mean of the local trait mix $\omega_i\mu_i(t) + (1-\omega_i)\mu_j(t)$ and the local optimum $\theta_i$ scales the recruitment rate for the population, and $\tilde{P}_i\sim {\rm Normal}(0,0.01)$ introduces a small amount of demographic variability.
The parameter $\tau$ is the strength of selection, and controls the sensitivity of recruitment to changes in the mean trait value away from the optimum.
Because straying individuals are emigrating from a population with a mean trait value farther from the local optimum, their rate of recruitment is diminished.
Recent studies of wild sockeye salmon have indeed found that straying individuals have lower life-time fitness than individuals that do not stray, although it is unknown at what life-stage this selection occurs \citep{Peterson:2014gy}.
% The compensatory effects are then determined by the exponential, following the Ricker stock-recruitment relationship.

### 2018-01-21 Manuscript/FinalDraft3/supp3.tex
labels to figure 4

We aim to understand how straying between populations may affect phenotypic evolution, and how in turn this evolutionary change may lead to changes in ecological dynamics as they unfold. 
To do so, we build upon Lande's classic formulation of phenotypic evolution \citep{Lande:1976ga}. In what follows, we  
1) re-derive Lande's phenotypic evolution equation for one population to introduce the reader to the main concepts and assumptions of the framework used, and,
2) explain how we used this framework for the case of two populations connected by dispersal.

@@

The model in the main text thus assumes the existence of a normally distributed trait in each population with means $\mu_{i}$ and $\mu_{j}$. Then straying occurs, and the the populations now have a mixture distribution as in \ref{12}. To keep track of the mean of that mix, we assume the mixture distribution to be Gaussian with mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$, which is true whenever straying is low enough, or the two means are not too different from one another. Then selection and reproduction occur at the same time leading to a new generation with a normally distributed trait distribution whose.

In a previous paper by Ronce and Kirkpatrick \citep{Ronce:2001dp}, it was shown that assumptions similar to the ones we make do not qualitatively alter the ecological or evolutionary dynamics observed.

@@

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.35\textwidth]{fig_density.png}
\caption{
Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
} \label{fig:symmetry}
\end{figure}

>>>>

We aim to understand how straying between populations may affect phenotypic evolution, and how this evolutionary change may in turn lead to changes in ecological dynamics as they unfold. 
To do so, we build upon Lande's classic formulation of phenotypic evolution \citep{Lande:1976ga}. In what follows, we  
1) re-derive Lande's phenotypic evolution equation for one population to introduce the reader to the main concepts and assumptions of the framework used, and,
2) explain how we used this framework for the case of two populations connected by dispersal.

@@

The model used in the main text thus assumes the existence of a normally distributed trait in each population with means $\mu_{i}$ and $\mu_{j}$. 
After straying occurs, the populations now have a mixture distribution as in \ref{12}. 
To keep track of the mean of that mix, we approximate the mixture distribution as a Gaussian with mean $\overline{x}=\omega_i \mu_{i}+(1-\omega_i)\mu_{j}$. 
Selection and reproduction thus occur at the same time leading to a new generation with a normally distributed trait distribution.
In a previous paper by Ronce and Kirkpatrick \citep{Ronce:2001dp}, it was shown that assumptions similar to those made here do not qualitatively alter the observed ecological or evolutionary dynamics.

@@

\begin{figure}
  \captionsetup{justification=raggedright,
singlelinecheck=false
}
\centering
\includegraphics[width=0.35\textwidth]{fig_density2.pdf}
\caption{
Steady-state densities of both populations as a function of $m$: one in a dominant state and one in a subordinate state.
Steady-states for populations with symmetrical values ($\alpha=0$) in the vital rates $r_{\rm max}$ and $\beta$ are shown with cool tones.
As the asymmetry among populations between sites increases ($\alpha>0$), their vital rates diverge, such that the maximal growth at sites 1 and 2 is now $r_{\rm max,1}=(1 + \alpha)r_{\rm max,2}$ and $\beta_1=(1+\alpha)\beta_2$ where $\alpha$ is increased from $0$ to $0.1$, thereby increasing the asymmetry in vital rates.
Steady-states for populations with increasingly asymmetric values ($\alpha\rightarrow 0.1$) for vital rates $r_{\rm max}$ and $\beta$ are shown in warmer tones.
The lower-density values that appear between $m=0.06$ and $m=0.1$ represent populations that have been trapped in the low-density basins of attraction associated with regime I.
Increasing the asymmetry in vital rates does not impact the qualitative nature of the dynamics.
} \label{fig:symmetry}
\end{figure}
